A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-19 podc_paper/podc-submission.tex
working on algorithm description

\newcommand{\here}[1]{\bf{[[#1]]}}
\newcommand{\ceil}[1]{\lceil #1 \rceil}
\newcommand{\nil}{nil}
\newcommand{\opa}[2]{\nf{#1(#2)}}
\newcommand{\op}[1]{\nf{#1}}
\newcommand{\var}[1]{\nf{#1}}

@@

\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\end{itemize}

@@

\begin{itemize}
  \item \tt{\sl{int} super}
  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} \size}
  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
\end{itemize}

@@

\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
\Function{Object}{Dequeue()}{} 
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \Return \tt{null} \label{returnNull}
\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
\While{\nf{start<end}}
\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
\If{\nf{blocks[mid].sum\sub{enq}<x}}
\State \nf{start= mid+1}
\Else
\State \nf{end= mid}
\EndIf
\EndWhile
\State\Return \nf{start}
\EndFunction{BinarySearch}

@@

\begin{algorithm}
\caption{\tt{\sl{Root}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{36}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}

@@

\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
\Function{void}{Propagate()}{}
\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
\State \Call{Refresh()}{} \label{secondRefresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}}
\State \tt{parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

@@

\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}

@@

\section{Proof of Correctness}

We adopt linearizability as our definition of correctness. In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue. The proof is structured like this. First, we define and prove some facts about blocks and the node's \nf{head} field. Then, we introduce  the linearization ordering formally. Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1). After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}. Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).

\subsection{Basic Properties}
In this subsection, we talk about some properties of blocks and fields of the tree nodes.

A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
\begin{definition}[Ordering of a block in a node]
Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
\end{definition}

Next, we show that the value of \nf{head} in a node can only be increased. By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
\begin{observation} \label{nonDecreasingHead}
For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
\end{observation}
\begin{proof}
  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
\end{proof}

\begin{lemma} \label{lem::headInc}
Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$). After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
\end{lemma}
\begin{proof}
If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds. Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}. By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
\end{proof}

Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
\end{invariant}
\begin{proof}
Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$. The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}. We show that if the invariant holds before such a change, then it still holds after the change.

In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}. Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail. Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.

The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}. If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}. \nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}. For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}. For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}. Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}. 
\end{proof}

We define the subblocks of a block recursively.

\begin{definition}[Subblock]\label{def::subblock}
A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.  We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
\end{definition}

The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
\begin{lemma} \label{lem::headProgress}
 If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
\end{lemma}
\begin{proof}
Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}. Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}. Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}. Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$. The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}. Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}. By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$. The proof for \nf{end\sub{right}} is similar.
\end{proof}

\begin{lemma} \label{lem::subblocksDistinct}
The sets of subblocks of any two blocks in a node are disjoint.  
\end{lemma}
\begin{proof}
We are going to prove the lemma by contradiction. Consider the lowest \nf{node} $n$ in the tree that violates the claim. Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$. Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap. Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$. By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$. Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
\end{proof}

\begin{definition}[Superblock]
  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
\end{definition}
\begin{corollary}
Every block has at most one superblock.
\end{corollary}
\begin{proof}
A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
\end{proof}

Now we can define the operations of a block using the definition of subblocks.
\begin{definition}[Operations of a block]\label{def::ops}
A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$. Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}. The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$. We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.  We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
\end{definition}

The next lemma proves that each operation appears at most once in the blocks of a node.
\begin{lemma}\label{lem::noDuplicates}
For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
\end{lemma}
\begin{proof}
We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}. Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$. From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$. Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks. This is a contradiction.
\end{proof}

\begin{definition}
$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$. An operation is \it{established} in node $n$ if it is in an established block of $n$. $EST^t_n$ is the set of established operations in node $n$ at time $t$.
\end{definition}

Now we want to say that blocks of a node grow over time.
\begin{observation}\label{lem::blocksOrder}
  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
\end{observation}
\begin{proof}
Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
\end{proof}

% \begin{corollary}\label{lem::establishedOrder}
%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
% \end{corollary}
% \begin{proof}
% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
% \end{proof}

\subsection{Ordering Operations}

\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/tree}
  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
\end{figure}

Now we define the ordering of operations stored in each node. In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately). Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right. We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
\end{definition}
\begin{itemize}
  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows. $E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation. If $n$ is an internal node, then
\begin{align*} 
E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows. $D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation. If $n$ is an internal node, then
\begin{align*} 
D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
\end{itemize}

The linearization ordering is given by the order in which operations appear in the blocks in the root.
\begin{definition}[Linearization] \label{def::lin}
 $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
\end{definition}

The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
\begin{observation}\label{sumToNum}
For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
\end{observation}

The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
\begin{lemma}\label{lem::numX}
Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.

\hspace{-\parindent}If $n$ is an internal node, then
\begin{enumerate}[label=(\arabic*)]
    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
\end{enumerate}
And for every node n, we have
\begin{enumerate}[label=(3)]
    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
\end{enumerate}
\end{lemma}
\begin{proof}
 We prove the claim by induction on the height of node $n$. For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true. Supposing the claim is true for $n$'s children, we prove the claim for $n$.
\begin{align*}
    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
\end{align*}
The first line follows from the Definition of \nf{num\sub{enq}}. The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine. Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3). (2) is similar to (1). Now we prove (3) starting from the definition of $E(n,b)$.
\begin{align*} 
E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
\end{proof}

>>>>

\newcommand{\here}[1]{\bf{[[#1]]}}
\newcommand{\ceil}[1]{\lceil #1 \rceil}
\newcommand{\nil}{nil}
\newcommand{\opa}[2]{\nf{#1(#2)}}
\newcommand{\op}[1]{\nf{#1}}
\newcommand{\var}[1]{\nf{#1}}
\newcommand{\com}{$\triangleright$}

@@

\begin{algorithm}
\caption{Objects used in Tournament Tree Data Structure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
\end{itemize}

@@

\begin{itemize}
  \item \tt{\sl{int} super}
  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{\com\ indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
\begin{itemize}
  \item \tt{\sl{int} \size}%
  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
\end{itemize}

@@

\Statex \com\ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
\Function{Object}{Dequeue()}{} 
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Statex \com\ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \Return \tt{null} \label{returnNull}
\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
\While{\nf{start<end}}
\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
\If{\nf{blocks[mid].sum\sub{enq}<x}}
\State \nf{start= mid+1}
\Else
\State \nf{end= mid}
\EndIf
\EndWhile
\State\Return \nf{start}
\EndFunction{BinarySearch}

@@

\begin{algorithm}
\caption{\tt{\sl{Root}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{36}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
\com\ the \nf{b}th block in the \nf{root}.}

@@

\Statex \com\ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
\Function{void}{Propagate()}{}
\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
\State \Call{Refresh()}{} \label{secondRefresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}}
\State \tt{parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

@@

\Statex \com\ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}

@@

\input{correctness}

\input{analysis}

\input{boundedspace}
